---
title: "Introduction to R"
author: "Connor Gilroy"
date: "`r Sys.Date()`"
output: 
  html_document:
    css: styles.css
  xaringan::moon_reader:
    css: [default, metropolis, metropolis-fonts]
    lib_dir: libs
    nature:
      highlightStyle: tomorrow
      highlightLines: true
      countIncrementalSlides: false
---

# What's R? What's RStudio?

**R** is a free, open-source statistical programming language. It's primarily used for statistics, data science, and other research. R has a vibrant community, including many people in the social sciences. 

**RStudio** is a program that makes it easier to work with R. It provides a nice environment for writing code and seeing all the output.

You can and should install both R and RStudio on your personal computer. 

---
# Tasks in statistical analysis

The basic things you'll want to do in R are the same as in Stata: 

- get data into the programming environment
- modify variables or create new ones, filter rows of observations
- look at variables and calculate descriptive statistics
- calculate inferential statistics
- visualize variables and models
- save output

---
# Familiar stuff

**Scripts.** Write and record all of your code in an R script (a `.R` file). This is like a Stata .do-file.

**The console.** Run your code line-by-line and see output by sending it to the console.

**Functions.** Functions are how R does things, like Stata commands. They take *arguments* (like Stata's varlist and options), and return something back. Arguments go inside parentheses, and might be referred to by name.

```{r eval=FALSE}
function_name(data, argument2 = 300, argument3 = FALSE, argument4 = "purple")
```

**Documentation.** Use `?` in the console to look up how something works, like typing `help` in Stata. For instance: `?mean`. 

---
# New stuff

**Things have names.** Stata can refer implicitly to whatever data you loaded or whatever model you fit. In R, you need to *assign* your data or model to a name, and refer to it by that name. 

```{r}
x <- 505
y <- c(504, 505, 506)
z <- "Social statistics"
```

**Add-ons are essential.** You can do most things with Stata's built-in commands. But many useful R tools live in *packages* that aren't installed or loaded by default. 

```{r eval=FALSE}
# install a package from an online repository
install.packages("package_name")

# load into your R session to use it
library(package_name)
```

---
# An example: loading data

**The goal:** take a file that's formatted for use in Stata ... and load it into R instead!

We're not the first humans to want to do this. So, some nice person on the internet has made a package called `haven` which promises to solve our problem. 

First, we install `haven`. We only do this once! (There should never be a line of code that says `install.packages` anywhere in your homework scripts.)

```{r eval=FALSE}
install.packages("haven")
```

---

Then, in every R session where we want to use `haven`, we have to load the package: 

```{r}
# load the package
library(haven)

# read the data into R, name it district_data
district_data <- read_dta("data/districtdata2020_compatible.dta")
```

For this to work, you need to tell R the right "path" to the data file. The place R starts looking for files is called the *working directory*. You can control where R looks in a number of ways.

---

```{r}
# look at the data
district_data
```


---
# Stuff to remember

**You're learning a language!** It takes time to get familiar and built habits. 

**You'll make lots of mistakes!** And that's okay! Errors are part of coding, and you'll get faster at fixing them. 

---
# What next?

I recommend two key resources: 

1. Chuck Lanfear's CSSS 508 course lectures + notes: https://clanfear.github.io/CSSS508/

2. The free online textbook *R for Data Science*: https://r4ds.had.co.nz/

**You should work through both of these when you have the time, starting now.**

*Want something easier?* The lab materials for Soc 225 are a gentle intro to exploratory data analysis in R: https://github.com/uw-soc-225/uw-soc-225

*Want something harder?* Advanced R is a free textbook about R as a programming language: https://adv-r.hadley.nz/
